# Model {.unnumbered}

```{r set}
#| message: false
#| warning: false
library(tidyverse)
```

We modelled the recovery of species richness over time using a model derived from the common framework for modelling the recovery of disturbed tropical forests [@maurent2023].

The recovery trajectory of species $richness$ over $time$ after disturbance is modelled using a long-term process LTP of recovery and a short-term process STP that allows for intermediate extremes (minimums or maximums) of species $richness$ at intermediate times. This results in the following equation:

$$\begin{align}
log(richness) \sim \mathcal N (log(\mu), \sigma)\\
\mu = \theta_{plot} \times (\phi_{plot} + LTP \times (1 - \phi_{plot}) + STP) \\
\end{align}$$

With $\theta_{plot}$ the plot asymptotic value of species richness recovery; $\phi_{plot}$ the plot remaining value of forest attribute relative to the plot asymptotic value just after disturbance (e.g. $1 - \phi_{plot}$ corresponds to disturbance relative intensity); $u$ the mean expected species richness in time after disturbance, and $\sigma$ the unexplained residual variation.

The long-term process $LTP$ recovery over $time$ after disturbance is modelled using an exponential recovery with a constant plot relative recovery rate $\lambda_{plot}$ as follow:

$$\begin{align}
LTP = 1 - e^{-\lambda_{plot} \times time} \\
\end{align}$$

The short-term process $STP$ that allows for intermediate extremes (minimums or maximums) of species richness at intermediate times is modelled using extremes timing parameters at the site level $\tau_{site}$ and intensity parameters at the plot level $\delta_{plot}$, as follows:

$$\begin{align}
STP = \delta_{plot} \times (\frac{time}{\tau_{site}} \times e^{1-\frac{time}{\tau_{site}}})^2 \\
\end{align}$$

We used hyperlaw normal distributions for parameters $\lambda_{plot}$ of plot relative recovery rate plot and $\tau_{site}$ extremes timing parameters at the site level site with means values $\mu_{\lambda}$ and $\mu_{\tau}$, and variances $\sigma_{\lambda}$ and $\sigma_{\tau}$. The full model with associated priors is further described in SI note (**To do**).

The relationship between disturbance intensity with observed basal area loss $BA~loss$ and the intermediate disturbance intensity parameters at the plot level $\delta_{plot}$ was tested *a posteriori* with a simple linear regression of $\delta_{plot}$ against $BA~loss$ across sites, to avoid any influence of $BA~loss$ on $\delta_{plot}$ inference.

```{r models}
#| message: false
#| warning: false
#| fig-cap: "Figure 2: Schematic representation of modelling of species richness over time following a selective logging disturbance. The x-axis represents the number of years post-disturbance. The green lines represent the modelled trajectory of either raw or rarefied species richness. The transparency of the lines represents the height of the intermediate diversity maximums, as described by the intermediate disturbance hypothesis below. The trajectory is modelled using the adapted common framework for modelling recovery in disturbed tropical forests (see the Methods section for details). Here, $delta_{plot}$ represents the diversity extremes (minimums or maximums) at the intermediate stage of the recovery trajectory. A positive $delta_{plot}$ directly represents the hypothesis H1 that an intermediate maximum of diversity will be observed over time. Similarly, positive relationship between $delta_{plot}$ and observed basal area loss directly represents the hypothesis H2 that the intermediate maximum of diversity is driven by increasing disturbance intensity."
ltp <- function(time, dist, thetainf, lambda) {
  theta0 <- thetainf * dist
  ltp <- 1 - exp(-lambda * time)
  theta0 + (thetainf - theta0) * (ltp)
}
stp <- function(time, dist, thetainf, lambda, tau, delta) {
  theta0 <- thetainf * dist
  ltp <- 1 - exp(-lambda * time)
  stp <- delta * (time / tau * exp(1 - time / tau))^2
  theta0 + ltp * (thetainf - theta0) + stp * thetainf
}
data.frame(time = seq(0.1, 30, length.out = 100)) |>
  mutate(y_stp_max = stp(time,
    dist = 0.5, thetainf = 20, lambda = 0.05,
    tau = 5, delta = 0.5
  )) |>
  mutate(y_stp_min = stp(time,
    dist = 0.5, thetainf = 20, lambda = 0.05,
    tau = 5, delta = 0.25
  )) |>
  mutate(y_ltp = ltp(time, dist = 0.5, thetainf = 20, lambda = 0.05)) |>
  bind_rows(data.frame(
    time = 0,
    y_stp_max = 20, y_stp_min = 20,
    y_ltp = 20,
    y_ba_max = 30, y_ba_min = 30
  )) |>
  ggplot(aes(time)) +
  geom_line(aes(y = y_ltp), col = "grey") +
  geom_line(aes(y = y_stp_max), col = "#a7d6ac") +
  geom_line(aes(y = y_stp_min), col = "#a7d6ac", alpha = .5) +
  theme_bw() +
  xlab("Years post-diturbance") +
  ylab("Species richness") +
  annotate("text",
    x = 4.5, y = 18, label = "delta[~plot]",
    col = "#a7d6ac", parse = TRUE, size = 5
  ) +
  annotate("text",
    x = 4, y = 14, label = "delta[~plot]",
    col = "#d2e9d4", parse = TRUE, size = 4
  ) +
  geom_segment(aes(x = 6, y = 13, xend = 6, yend = 22),
    arrow = arrow(length = unit(0.2, "cm")), col = "#a7d6ac"
  ) +
  geom_segment(aes(x = 6, y = 22, xend = 6, yend = 13),
    arrow = arrow(length = unit(0.2, "cm")), col = "#a7d6ac"
  ) +
  geom_segment(aes(x = 5, y = 12.5, xend = 5, yend = 17),
    arrow = arrow(length = unit(0.2, "cm")), col = "#d2e9d4"
  ) +
  geom_segment(aes(x = 5, y = 17, xend = 5, yend = 12.5),
    arrow = arrow(length = unit(0.2, "cm")), col = "#d2e9d4"
  ) +
  geom_point(x = 5.5, y = 22.3, col = "#a7d6ac") +
  xlab("Years post-diturbance")
```

## Fits

Chunks below help running the models for raw and rarefied species richness.

```{r fit_raw}
#| eval: false
library(cmdstanr)
out <- file.path("chains", "raw")
if (file.exists(out)) {
  unlink(out, recursive = TRUE)
}
dir.create(out)
model <- cmdstan_model("models/model.stan")
data <- read_tsv("data/derived_data/data.tsv")
ind <- data |>
  select(site, sitenum, plot, plotnum, dist_obs) %>%
  unique()
data_u <- ind |>
  left_join(filter(data, rel_year <= 0))
data_r <- ind |>
  left_join(filter(data, rel_year > 0))
mdata <- list(
  n = nrow(data_r),
  s = max(data_r$sitenum),
  p = max(data_r$plotnum),
  y = data_r$diversity_q0_sp,
  t = data_r$rel_year,
  site = data_r$sitenum,
  plot = data_r$plotnum,
  site_plot = ind$sitenum,
  n_u = nrow(data_u),
  y_u = data_u$diversity_q0_sp,
  plot_u = data_u$plotnum,
  dist_p = ind$dist_obs
)
fit <- model$sample(
  mdata,
  chains = 4,
  parallel_chains = 4,
  save_warmup = FALSE,
  refresh = 100,
  output_dir = out
)
```

```{r fit_rarefied}
#| eval: false
library(cmdstanr)
out <- file.path("chains", "rarefied")
if (file.exists(out)) {
  unlink(out, recursive = TRUE)
}
dir.create(out)
model <- cmdstan_model("models/model.stan")
data <- read_tsv("data/derived_data/data.tsv")
ind <- data |>
  select(site, sitenum, plot, plotnum, dist_obs) %>%
  unique()
data_u <- ind |>
  left_join(filter(data, rel_year <= 0))
data_r <- ind |>
  left_join(filter(data, rel_year > 0))
mdata <- list(
  n = nrow(data_r),
  s = max(data_r$sitenum),
  p = max(data_r$plotnum),
  y = data_r$diversity_q0_sp_rar,
  t = data_r$rel_year,
  site = data_r$sitenum,
  plot = data_r$plotnum,
  site_plot = ind$sitenum,
  n_u = nrow(data_u),
  y_u = data_u$diversity_q0_sp_rar,
  plot_u = data_u$plotnum,
  dist_p = ind$dist_obs
)
fit <- model$sample(
  mdata,
  chains = 4,
  parallel_chains = 4,
  save_warmup = FALSE,
  refresh = 100,
  output_dir = out
)
```
